### 2019-11-27 ic-journal/part2.tex
Manually merged and synced changes in part2.tex

\subsubsection{Weak bisimilarity}

Milner's book~\cite{Mil89} contains only two ``unique solutions of
equations'' theorems, one for strong equivalence, the other for
observational congruence.   There is however a
version  of the result for weak bisimilarity that shares a large portion of proof
steps with the proof for observational congruence.  
As  weak bisimilarity 
 is not \hl{a} congruence, we have to be more restrictive on the syntax for
 the equation, using strong guardedness.
% to make correct statement of this
% theorem, we must slightly modify the concept of sequential expresses
% with further restrictions: no direct sums (or only guarded sums).
% The ``unique solution of equations'' theorem for weak equivalence
% requires more restrictions on the expression, namely, .
% Now we're ready to state and prove the following lemma:
\begin{lemma}[\texttt{WEAK_UNIQUE_SOLUTION_LEMMA}]
If the variable $X$ is guarded and sequential in $G$, and
$G\{P/X\}\overset{\alpha}{\rightarrow} P'$, then $P'$ takes the form
$H\{P/X\}$, for some expression $H$, and for any $Q$, we have
$G\{Q/X\}\overset{\alpha}{\rightarrow} H\{Q/X\}$. Moreover $H$ is
sequential, and if $\alpha = \tau$, then $H$ is also guarded.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{SG} \HOLFreeVar{G} \HOLSymConst{\HOLTokenConj{}} \HOLConst{GSEQ} \HOLFreeVar{G} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{a} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLFreeVar{G} \HOLBoundVar{P} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{H}. \HOLConst{GSEQ} \HOLBoundVar{H} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLBoundVar{a} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{SG} \HOLBoundVar{H}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \HOLBoundVar{H} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLFreeVar{G} \HOLBoundVar{Q} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \HOLBoundVar{H} \HOLBoundVar{Q}
\end{alltt}
\end{lemma}

\subsubsection{Rooted bisimilarity}

@@

This lemma plays a key role in the
corresponding `unique solutions' theorem, in connection with the
following result. 
In this part, we have not 
used 
 ``bisimulation up-to'' techniques. Using
 the  above \texttt{OBS_CONGR_BY_WEAK_BISIM},
 we  directly construct the appropriate bisimulation relations.
\begin{lemma}[\texttt{OBS_UNIQUE_SOLUTION_LEMMA}]
If the variable $X$ is guarded and sequential in $G$, and
$G\{P/X\}\overset{\alpha}{\rightarrow} P'$, then $P'$ takes the form
$H\{P/X\}$, for some  $H$, and for any $Q$, we also have 
$G\{Q/X\}\overset{\alpha}{\rightarrow} H\{Q/X\}$. Moreover $H$ is
sequential, and if $\alpha = \tau$ then $H$ is also guarded.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{SG} \HOLFreeVar{G} \HOLSymConst{\HOLTokenConj{}} \HOLConst{SEQ} \HOLFreeVar{G} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{a} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLFreeVar{G} \HOLBoundVar{P} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{H}. \HOLConst{SEQ} \HOLBoundVar{H} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLBoundVar{a} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{SG} \HOLBoundVar{H}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \HOLBoundVar{H} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLFreeVar{G} \HOLBoundVar{Q} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \HOLBoundVar{H} \HOLBoundVar{Q}
\end{alltt}
\end{lemma}

@@

@@

The definition of ``unique label'' can be done in many \hl{ways. The}
following definition (a suggestion from Robert Beers)
avoids any counting or filtering in the list.
It says that a label is unique in a list of actions if and only if there is no
label in the rest of list:
\begin{alltt}
   \HOLConst{UNIQUE_LABEL} \HOLFreeVar{u} \HOLFreeVar{L} \HOLTokenDefEquality{}
     \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{L}\sb{\mathrm{1}}} \ensuremath{\HOLBoundVar{L}\sb{\mathrm{2}}}. \ensuremath{\HOLBoundVar{L}\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenDoublePlus} \ensuremath{[}\HOLFreeVar{u}\ensuremath{]} \HOLSymConst{\HOLTokenDoublePlus} \ensuremath{\HOLBoundVar{L}\sb{\mathrm{2}}} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{L} \HOLSymConst{\HOLTokenConj{}} \HOLConst{NO_LABEL} \ensuremath{\HOLBoundVar{L}\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{NO_LABEL} \ensuremath{\HOLBoundVar{L}\sb{\mathrm{2}}}\hfill{[UNIQUE_LABEL_def]}
\end{alltt}

@@

Traces are actually used in the proof of above lemma via 
the following lemma (\texttt{unfolding_lemma4}):
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{GCONTEXT} \HOLFreeVar{C} \HOLSymConst{\HOLTokenConj{}} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{TRACE} \ensuremath{(}\ensuremath{(}\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}\ensuremath{)} \HOLFreeVar{P}\ensuremath{)} \HOLFreeVar{xs} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{xs} \HOLSymConst{\HOLTokenLeq{}} \HOLFreeVar{n} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{C}\sp{\prime}}. \HOLConst{GCONTEXT} \ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLConst{TRACE} \ensuremath{(}\ensuremath{(}\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}\ensuremath{)} \HOLBoundVar{Q}\ensuremath{)} \HOLFreeVar{xs} \ensuremath{(}\ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLBoundVar{Q}\ensuremath{)}
\end{alltt}
which roughly says that for any context $C$ and weakly-guarded context
$E$, if $C [\, E^n[P]\,] \overset{xs}{\Longrightarrow} P'$ and the length
of actions $xs \leqslant n$, then \hl{$P'$} has the form of $C'[P]$.
Traces are used \hl{in} reasoning about the number of intermediate actions in weak
transitions. For instance, from Def.~\ref{d:BisCon}, it is easy
to see that a weak transition either becomes shorter
or remains the same when moving between $\mcontrBIS$-related processes.
This property is essential in the proof of
Lemma~\ref{l:uptocon}. We show only one such lemma, for the case of
visible weak transitions:
% passing into $\mcontrBIS$ (from left to right):
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{xs} \HOLBoundVar{l} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLConst{TRACE} \HOLFreeVar{P} \HOLBoundVar{xs} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{UNIQUE_LABEL} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \HOLBoundVar{xs} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{xs}\sp{\prime}} \ensuremath{\HOLBoundVar{Q}\sp{\prime}}.
           \HOLConst{TRACE} \HOLFreeVar{Q} \ensuremath{\HOLBoundVar{xs}\sp{\prime}} \ensuremath{\HOLBoundVar{Q}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \ensuremath{\HOLBoundVar{xs}\sp{\prime}} \HOLSymConst{\HOLTokenLeq{}} \HOLConst{LENGTH} \HOLBoundVar{xs} \HOLSymConst{\HOLTokenConj{}}
           \HOLConst{UNIQUE_LABEL} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \ensuremath{\HOLBoundVar{xs}\sp{\prime}}\hfill{[contracts_AND_TRACE_label]}
\end{alltt}

>>>>

\subsubsection{The case of rooted bisimilarity}

@@

This lemma plays a key role in the
corresponding `unique solutions' theorem, in connection with the
following result. 
In this part, we have not 
used 
 ``bisimulation up-to'' techniques. Using
 the  above \texttt{OBS_CONGR_BY_WEAK_BISIM},
 we  directly construct the appropriate bisimulation relations.
\begin{lemma}[\texttt{OBS_UNIQUE_SOLUTION_LEMMA}]
If the variable $X$ is guarded and sequential in $G$, and
$G\{P/X\}\overset{\alpha}{\rightarrow} P'$, then $P'$ takes the form
$H\{P/X\}$, for some  $H$, and for any $Q$, we also have 
$G\{Q/X\}\overset{\alpha}{\rightarrow} H\{Q/X\}$. Moreover $H$ is
sequential, and if $\alpha = \tau$, then $H$ is also guarded.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{SG} \HOLFreeVar{G} \HOLSymConst{\HOLTokenConj{}} \HOLConst{SEQ} \HOLFreeVar{G} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{a} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLFreeVar{G} \HOLBoundVar{P} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{H}. \HOLConst{SEQ} \HOLBoundVar{H} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLBoundVar{a} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{SG} \HOLBoundVar{H}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \HOLBoundVar{H} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLFreeVar{G} \HOLBoundVar{Q} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \HOLBoundVar{H} \HOLBoundVar{Q}
\end{alltt}
\end{lemma}

@@

\subsubsection{The case of weak bisimilarity}

Milner~\cite{Mil89} \emph{only mentioned} two ``unique solution of
equations'' theorems, one for strong equivalence, the other for
rooted bisimilarity.   There is, however, another
version for weak bisimilarity that shares a large portion of proof
steps with the proof for \hl{rooted bisimilarity.}
As weak bisimilarity is not a congruence, we have to be more restrictive
on the syntax of the equation, using strong guardedness.

\begin{lemma}[\texttt{WEAK_UNIQUE_SOLUTION_LEMMA}]
If the variable $X$ is guarded and sequential in $G$, and
$G\{P/X\}\overset{\alpha}{\rightarrow} P'$, then $P'$ takes the form
$H\{P/X\}$, for some expression $H$, and for any $Q$, we have
$G\{Q/X\}\overset{\alpha}{\rightarrow} H\{Q/X\}$. Moreover $H$ is
sequential, and if $\alpha = \tau$ then $H$ is also guarded.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{SG} \HOLFreeVar{G} \HOLSymConst{\HOLTokenConj{}} \HOLConst{GSEQ} \HOLFreeVar{G} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{a} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLFreeVar{G} \HOLBoundVar{P} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{H}. \HOLConst{GSEQ} \HOLBoundVar{H} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLBoundVar{a} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\ensuremath{\tau}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{SG} \HOLBoundVar{H}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \HOLBoundVar{H} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLFreeVar{G} \HOLBoundVar{Q} \HOLTokenTransBegin\HOLBoundVar{a}\HOLTokenTransEnd \HOLBoundVar{H} \HOLBoundVar{Q}
\end{alltt}
\end{lemma}

@@

The definition of ``unique label'' can be done in many ways. The
following definition (a suggestion from Robert Beers)
avoids any counting or filtering in the list.
It says that a label is unique in a list of actions if and only if there is no
label in the rest of list:
\begin{alltt}
   \HOLConst{UNIQUE_LABEL} \HOLFreeVar{u} \HOLFreeVar{L} \HOLTokenDefEquality{}
     \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{L}\sb{\mathrm{1}}} \ensuremath{\HOLBoundVar{L}\sb{\mathrm{2}}}. \ensuremath{\HOLBoundVar{L}\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenDoublePlus} \ensuremath{[}\HOLFreeVar{u}\ensuremath{]} \HOLSymConst{\HOLTokenDoublePlus} \ensuremath{\HOLBoundVar{L}\sb{\mathrm{2}}} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{L} \HOLSymConst{\HOLTokenConj{}} \HOLConst{NO_LABEL} \ensuremath{\HOLBoundVar{L}\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{NO_LABEL} \ensuremath{\HOLBoundVar{L}\sb{\mathrm{2}}}\hfill{[UNIQUE_LABEL_def]}
\end{alltt}

@@

Traces are actually used in the proof of above lemma via 
the following lemma (\texttt{unfolding_lemma4}):
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{GCONTEXT} \HOLFreeVar{C} \HOLSymConst{\HOLTokenConj{}} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{TRACE} \ensuremath{(}\ensuremath{(}\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}\ensuremath{)} \HOLFreeVar{P}\ensuremath{)} \HOLFreeVar{xs} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{xs} \HOLSymConst{\HOLTokenLeq{}} \HOLFreeVar{n} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{C}\sp{\prime}}. \HOLConst{GCONTEXT} \ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \ensuremath{\HOLFreeVar{P}\sp{\prime}} \HOLSymConst{\ensuremath{=}} \ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Q}. \HOLConst{TRACE} \ensuremath{(}\ensuremath{(}\HOLFreeVar{C} \HOLSymConst{\HOLTokenCompose} \HOLConst{FUNPOW} \HOLFreeVar{E} \HOLFreeVar{n}\ensuremath{)} \HOLBoundVar{Q}\ensuremath{)} \HOLFreeVar{xs} \ensuremath{(}\ensuremath{\HOLBoundVar{C}\sp{\prime}} \HOLBoundVar{Q}\ensuremath{)}
\end{alltt}
which roughly says that for any context $C$ and weakly-guarded context
$E$, if $C [\, E^n[P]\,] \overset{xs}{\Longrightarrow} P'$ and the length
of actions $xs \leqslant n$, then $P'$ has the form of $C'[P]$.
Traces are used in reasoning about the number of intermediate actions in weak
transitions. For instance, from Def.~\ref{d:BisCon}, it is easy
to see that, a weak transition either becomes shorter
or remains the same when moving between $\mcontrBIS$-related processes.
This property is essential in the proof of
Lemma~\ref{l:uptocon}. We show only one such lemma, for the case of
visible weak transitions:
%passing into $\mcontrBIS$ (from left to right):
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{xs} \HOLBoundVar{l} \ensuremath{\HOLBoundVar{P}\sp{\prime}}.
       \HOLConst{TRACE} \HOLFreeVar{P} \HOLBoundVar{xs} \ensuremath{\HOLBoundVar{P}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{UNIQUE_LABEL} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \HOLBoundVar{xs} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\ensuremath{\HOLBoundVar{xs}\sp{\prime}} \ensuremath{\HOLBoundVar{Q}\sp{\prime}}.
           \HOLConst{TRACE} \HOLFreeVar{Q} \ensuremath{\HOLBoundVar{xs}\sp{\prime}} \ensuremath{\HOLBoundVar{Q}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLFreeVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{Q} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \ensuremath{\HOLBoundVar{xs}\sp{\prime}} \HOLSymConst{\HOLTokenLeq{}} \HOLConst{LENGTH} \HOLBoundVar{xs} \HOLSymConst{\HOLTokenConj{}}
           \HOLConst{UNIQUE_LABEL} \ensuremath{(}\HOLConst{label} \HOLBoundVar{l}\ensuremath{)} \ensuremath{\HOLBoundVar{xs}\sp{\prime}}\hfill{[contracts_AND_TRACE_label]}
\end{alltt}
